---
title: 'Reproducible Research: Peer Assessment 1'
author: "Simon Coburg"
date: "26/02/2021"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*This is my submission for the Reproducible Research Course Project 1. To obtain more information about this project see the [ReadMe](https://github.com/NestaKobe/RepData_PeerAssessment1/blob/master/README.md) on GitHub*.

## Loading and preprocessing the data
The following steps were applied to load, import and check the dataset.

```{r presettings, loading data}
#Libraries
library(ggplot2)
library(dplyr)


#Load data
path <- getwd()
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
              , destfile = paste(path, "dataFiles.zip", sep = "/"))
unzip(zipfile = "dataFiles.zip")

#Check if data present
dir() 

#Remove zip file
file.remove("dataFiles.zip")

#Import dataset
activity <- read.csv("activity.csv")

#Check dataset
head(activity)
str(activity)
```


## What is mean total number of steps taken per day?
For this part of the assignment, missing values in the dataset could be ignored.

#### 1. Number of steps taken per day
```{r}
steps_total <- activity %>%
  group_by(date) %>%
  summarise(steps_daily = sum(steps, na.rm = TRUE))
head(steps_total)

#Calculate and report sum
sum(steps_total$steps_daily, na.rm = TRUE)

```
Total number of steps taken are *570,608*.

#### 2. Create histogram of total number of steps taken each day
```{r}
#Create barplot
ggplot(steps_total, aes(steps_daily)) + 
      geom_histogram(binwidth = 2500) +
      geom_rug(aes(steps_daily)) +
      ggtitle("Total number of steps taken each day") +
      xlab("Steps") + 
      ylab("Frequency") +
      scale_y_continuous(breaks=seq(0,18,2)) +
      scale_x_continuous(breaks=seq(0,25000,2500)) +
      theme(plot.title = element_text(hjust = 0.5))
```

#### 3. Calculate mean and median of total number of steps taken per day

```{r}
summary(steps_total$steps_daily, na.rm = TRUE)
```


## What is the average daily activity pattern?

#### 1. Time series of the averaged number of steps taken by 5-minute intervals
```{r}

#Create line plot for average steps per 5-minute intervals
steps_interval <- activity %>% 
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm =TRUE))


ggplot(steps_interval, aes(interval, steps)) +
      geom_line() +
      ggtitle("Average daily activity pattern") +
      xlab("5-minute intervals") +
      ylab("Average number of steps taken") +
      theme(plot.title = element_text(hjust = 0.5))
```
#### 2. Which 5-minute interval (averaged across all days) contains max. number of steps?

```{r}
#Report max 5-minute interval
which.max(steps_interval$steps)
max_interval = round(steps_interval[104,])
max_interval
```
The 5-minute interval that contains the maximum number of steps is the *835th interval*.

## Imputing missing values

#### 1. Calculate total number of NA in the dataset
```{r}
#Creating table
missing <- tbl_df(activity)
#Filtering for NA
missing %>% filter(is.na(steps)) %>% summarize(missing_values = n())
```

#### 2. Replacing NA values by the averaged 5-minute intervals
```{r}
activity$steps_complete <- ifelse(is.na(activity$steps), 
  round(steps_interval$steps[match(activity$interval, steps_interval$interval)],0), 
  activity$steps)
```

#### 3. Create new dataset adjusted for NA values
```{r}
activity_complete <- data.frame(steps=activity$steps_complete, 
  interval=activity$interval, date=activity$date)

head(activity_complete)

#Check if NA values still present
any(is.na(activity_complete))
```

#### 4. Create histogram with new dataset
```{r}
activity_complete_steps <- aggregate(activity_complete$steps, list(activity_complete$date), FUN=sum)
colnames(activity_complete_steps) <- c("Date", "Steps")
head(activity_complete_steps)


ggplot(activity_complete_steps, aes(Steps)) +
      geom_histogram(binwidth = 2500) +
      ggtitle("Adjusted total number of steps taken each day") +
      xlab("Steps") + 
      ylab("Frequency") +
      theme(plot.title = element_text(hjust = 0.5))
```

#### 5. Calculate and report sum, mean & median
```{r}
sum(activity_complete_steps$Steps, na.rm = TRUE)
summary(activity_complete_steps$Steps, na.rm = TRUE)

#Do sum/mean/median values differ from Q1
sum(activity_complete_steps$Steps, na.rm = TRUE) - sum(steps_total$steps_daily, na.rm = TRUE)
summary(activity_complete_steps$Steps, na.rm = TRUE) - summary(steps_total$steps_daily, na.rm = TRUE)

```
The estimate of total number of steps increases by *86,096*. Mean and median differ by the amount of *1,411* and *367* respectively compared to the previous estimates that excluded the missing values.

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable with two levels: "weekday" and "weekend"
```{r}
#Change date format (if necessary)
activity_complete$date <- as.Date(activity_complete$date, format = "%Y-%m-%d")

#Add variable with the according weekdays name
activity_complete$Weekday <- weekdays(activity_complete$date)

#Distinguish between weekdays and weekend
activity_complete$Type <- ifelse(activity_complete$Weekday=='Saturday' | 
  activity_complete$Weekday=='Sunday', 'Weekend','Weekday')

#Check dataset
head(activity_complete)
```

#### 2. Plot two time series for weekdays & weekend
```{r}
steps_week <- activity_complete %>% 
  group_by(interval, Type) %>%
  summarise(steps = mean(steps, na.rm =TRUE))

ggplot(steps_week, aes(interval, steps)) +
      geom_line() +
      facet_wrap(~Type, nrow=2) +
      ggtitle("Weekday vs weekend activity pattern") +
      xlab("5-minute intervals") +
      ylab("Average number of steps taken") +
      theme_bw() +
      theme(plot.title = element_text(hjust = 0.5))
```

